---
title: "Final code"
author: "Ziliang Song"
date: "2026-05-05"
output: html_document
---





##EDA part
```{r}
training<- read.csv("Completed_train.csv")

testing<- read.csv("Completed_test.csv")
#Check the structure of both datasets.
str(training)
str(testing)
```

#Convert the completion_date and start_date from character to date format, and examine their distribution and basic statistics
```{r}
training <- training %>%
  mutate(
    start_date = as.Date(start_date),
    completion_date = as.Date(completion_date)
  )
testing <- testing %>%
  mutate(
    start_date = as.Date(start_date),
    completion_date = as.Date(completion_date)
  )
summary(training$completion_date)
summary(testing$completion_date)
```
#A quick overview of duration distribution
```{r}
train_test <- bind_rows(training, testing)

summary(train_test$time_day)

sd(train_test$time_day, na.rm = TRUE)
```

##TARGET VARIABLE
```{r}

ggplot(train_test, aes(x = time_day)) +
  geom_histogram(
    bins = 50,
    fill = "#6a5acd",
    alpha = 0.7,
    color = "white"
  ) +
  geom_vline(
    aes(xintercept = mean(time_day, na.rm = TRUE)),
    color = "red",
    linetype = "dashed",
    size = 1
  ) +
  geom_vline(
    aes(xintercept = median(time_day, na.rm = TRUE)),
    color = "blue",
    linetype = "dashed",
    size = 1
  ) +
  labs(
    title = "Trial Duration Distribution",
    subtitle = "Red = Mean, Blue = Median",
    x = "Duration (Days)",
    y = "Count"
  ) +
  theme_minimal(base_size = 14)
```

#CATEGORICAL IMPACT
```{r}
ggplot(train_test, aes(x = phase.x)) +
  geom_bar(
    fill = "#72B7B2",
    alpha = 0.9
  ) +
  stat_count(
    geom = "text",
    aes(label = after_stat(count)),
    vjust = -0.3,
    size = 4
  ) +
  labs(
    title = "Distribution of Trial Phases",
    x = "Phase",
    y = "Count"
  ) +
  theme_minimal(base_size = 14)

```


```{r}
ggplot(train_test, aes(x = phase.x, y = time_day, fill = phase.x)) +
  geom_boxplot(
    width = 0.6,
    outlier.alpha = 0.2,
    outlier.size = 1
  ) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Trial Duration Variance by Phase",
    subtitle = "Distribution of duration across different trial phases",
    x = "Trial Phase",
    y = "Duration (Days)"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "none",
    plot.title = element_text(face = "bold", size = 18),
    plot.subtitle = element_text(size = 13, color = "gray40"),
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12)
  )

```



#SIZE CORRELATION
```{r}
ggplot(train_test, aes(x = enrollment)) +
  geom_histogram(
    binwidth = 50, 
    fill = "#4C78A8",
    color = "white",
    alpha = 0.9
  ) +
  coord_cartesian(xlim = c(0, 2000)) + 
  labs(
    title = "Enrollment Distribution",
    subtitle = "Bin width = 50",
    x = "Enrollment",
    y = "Count"
  ) +
  theme_minimal(base_size = 14)
```

#Relationship between enrollment and duration
```{r}
ggplot(train_test %>% filter(enrollment > 0), aes(x = enrollment, y = time_day)) +
  geom_point(
    alpha = 0.2,
    size = 1,
    color = "#2C5AA0"
  ) +
  geom_smooth(
    method = "loess",
    se = FALSE,
    color = "#D55E00",
    linewidth = 1.2
  ) +
  scale_x_log10() +
  theme_minimal(base_size = 14) +
  labs(
    title = "Enrollment vs. Trial Duration",
    subtitle = "Enrollment shown on a log10 scale",
    x = "Enrollment (log10 scale)",
    y = "Duration (Days)"
  )


```

#TEXT LENGTH
```{r}
ziliang_data %>%
  mutate(word_count = str_count(criteria, "\\w+")) %>%
  ggplot(aes(x = word_count)) +
  geom_histogram(
    bins = 50,
    fill = "#2A9D8F",
    color = "white",
    alpha = 0.9
  ) +
  geom_vline(
    xintercept = 512,
    linetype = "dashed",
    color = "#E76F51",
    linewidth = 1.2
  ) +
  annotate(
    "text",
    x = 512,
    y = Inf,                   
    label = "512 Token Limit",
    color = "#E76F51",
    vjust = 1.5,
    hjust = 1.1,
    size = 4
  ) +
  coord_cartesian(xlim = c(0, 2000)) + 
  theme_minimal(base_size = 14) +
  labs(
    title = "Distribution of Criteria Text Length",
    subtitle = "Word count used as a proxy for token length",
    x = "Word Count",
    y = "Count"
  ) +
  theme(
    plot.title = element_text(face = "bold", size = 18),
    plot.subtitle = element_text(size = 12, color = "gray40"),
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 14)
  )
```

#outlier detection
```{r}
Q1 <- quantile(train_test$time_day, 0.25, na.rm = TRUE)
Q3 <- quantile(train_test$time_day, 0.75, na.rm = TRUE)
IQR_val <- Q3 - Q1

lower_bound <- Q1 - 1.5 * IQR_val
upper_bound <- Q3 + 1.5 * IQR_val

train_test <- train_test %>%
  mutate(
    is_outlier = time_day > upper_bound | time_day < lower_bound
  )

outlier_ratio <- mean(train_test$is_outlier, na.rm = TRUE)
print(outlier_ratio)


ggplot(train_test, aes(y = time_day)) +
  geom_jitter(
    aes(x = 0, color = is_outlier),
    alpha = 0.2,
    width = 0.15,
    size = 0.8
  ) +
  geom_boxplot(
    width = 0.3,
    fill = "orange",
    color = "black",
    alpha = 0.6,
    outlier.shape = NA   
  ) +
  geom_hline(
    yintercept = upper_bound,
    linetype = "dashed",
    color = "red",
    linewidth = 1
  ) +
  annotate(
    "text",
    x = -0.1,
    y = upper_bound,
    label = paste0("Cutoff = ", round(upper_bound, 1)),
    color = "blue",
    hjust = 0,
    vjust = -1,
    size = 4
  )+
  scale_color_manual(values = c("FALSE" = "grey", "TRUE" = "#E63946")) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "none",
    axis.title.x = element_blank(),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank()
  ) +
  labs(
    title = "Outlier Detection for Trial Duration",
    subtitle = "Based on 1.5 × IQR rule",
    y = "Duration (Days)"
  )
```






##Ablation
```{r}
library(readr)
library(dplyr)
library(stringr)
library(purrr)
library(ggplot2)
```


```{r}
folder_path <- "Ablation_result"

file_list <- list.files(
  path = folder_path,
  pattern = "\\.csv$",
  full.names = TRUE
)

data_list <- lapply(file_list, function(file) {
  read_csv(file, show_col_types = FALSE)
})

names(data_list) <- basename(file_list)

prediction_files <- names(data_list)[
  str_detect(names(data_list), "_predictions\\.csv$")
]

prediction_list <- data_list[prediction_files]

length(prediction_list)
names(prediction_list)

all_pred <- map_dfr(prediction_files, function(file_name) {
  
    df <- prediction_list[[file_name]]
  
  base_name <- str_remove(file_name, "\\.csv$")
  base_name <- str_remove(base_name, "_predictions$")
  
  parts <- str_split(base_name, "__", simplify = TRUE)
  
  df %>%
    mutate(
      source_file = file_name,
      feature_setting = parts[1],
      model = parts[2]
    )
})

metrics_df <- all_pred %>%
  group_by(model, feature_setting) %>%
  summarise(
    n = n(),
    MAE = mean(abs(y_true - y_pred), na.rm = TRUE),
    RMSE = sqrt(mean((y_true - y_pred)^2, na.rm = TRUE)),
    R2 = 1 - sum((y_true - y_pred)^2, na.rm = TRUE) / 
      sum((y_true - mean(y_true, na.rm = TRUE))^2, na.rm = TRUE),
    Spearman = cor(y_true, y_pred, method = "spearman", use = "complete.obs"),
      .groups = "drop"
  )

metrics_df%>%
  arrange(MAE)
```


##Histogram of Abaltion
```{r}
library(viridis)

plot_df <- metrics_df %>%
  mutate(label = paste(model, feature_setting, sep = " | "))

ggplot(plot_df, aes(y = reorder(label, MAE), x = MAE, fill = MAE)) +
  geom_col() +
  scale_fill_viridis(option = "C") +
  labs(
    title = "MAE by Model and Feature Setting",
    y = "Model | Feature Setting",
    x = "MAE"
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.text.y = element_text(size = 8)
  )

ggplot(plot_df, aes(y = reorder(label, R2), x = R2, fill = R2)) +
  geom_col() +
  scale_fill_viridis(option = "D") +
  labs(
    title = "R² by Model and Feature Setting",
    y = "Model | Feature Setting",
    x = "R²"
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.text.y = element_text(size = 8)
  )

```







##Metrics by phase
```{r}
df <- read_csv("final_phase_metrics.csv")

df_all <- df %>%
  filter(experiment == "all_features")

models <- unique(df_all$model)

for (m in models) {
  
  cat("\n====================================\n")
  cat("Model:", m, "\n")
  cat("====================================\n")
  
  tab <- df_all %>%
    filter(model == m) %>%
    mutate(
      phase = as.character(phase),
      phase = factor(phase, levels = c("All", "1", "2", "3", "4"))
    ) %>%
    arrange(phase) %>%
    transmute(
      Phase = phase,
      `MAE (↓)` = sprintf("%.3f", MAE),
      `RMSE (↓)` = sprintf("%.3f", RMSE),
      `R^2 (↑)` = sprintf("%.3f", R2),
      `Pearson Corr. (↑)` = sprintf("%.3f", Pearson)
    )
  
  print(tab)
}
```

##Reuslts of Machine learning Methods
```{r}
baseline <- read_csv("all_model_results.csv")

baseline <- baseline %>%
  mutate(model = as.character(model))

transformer <- read_csv("transformer_mlp_metrics.csv")

transformer <- transformer %>%
  mutate(model = "Transformer + MLP")

common_cols <- intersect(names(baseline), names(transformer))

baseline2 <- baseline %>% select(all_of(common_cols))
transformer2 <- transformer %>% select(all_of(common_cols))

compare_df <- bind_rows(baseline2, transformer2)

final_table <- compare_df %>%
  select(model, MAE, MSE, RMSE, R2, Pearson_r) %>%
  mutate(
    MAE = round(MAE, 3),
    MSE = round(MSE, 3),
    RMSE = round(RMSE, 3),
    R2 = round(R2, 3),
    Pearson_r = round(Pearson_r, 3)
  )

print(final_table)
```


##Hyperparameter setting
```{r}
hyp <- read_csv("hyperparameter_search_summary.csv")
table_hyp <- hyp %>%
  select(
    name,
    d_model,
    nhead,
    num_layers,
    dim_feedforward,
    dropout,
    mlp_hidden,
    lr,
    weight_decay,
    MAE,
    RMSE,
    R2,
    Pearson_r
  ) %>%
  arrange(RMSE)

print(table_hyp)
table_hyp <- hyp %>%
  select(name, MAE, RMSE, R2, Pearson_r) %>%
  arrange(RMSE)

print(table_hyp)
```

##Distribution of Duration
```{r}
training <- read_csv(
 "time_prediction_test_filtered_with_phase_enrollment.csv"
)

testing <- read_csv(
  "time_prediction_train_filtered_with_phase_enrollment.csv"
)

train_filtered <- train_filtered %>%
  mutate(phase = coalesce(phase.y, phase.x)) %>%
  select(-phase.x, -phase.y)


ggplot(train_filtered, aes(x = time_day)) +
  geom_histogram(
    bins = 50,
    fill = "#6a5acd",
    alpha = 0.7,
    color = "white"
  ) +
  geom_vline(
    aes(xintercept = mean(time_day, na.rm = TRUE)),
    color = "red",
    linetype = "dashed",
    size = 1
  ) +
  geom_vline(
    aes(xintercept = median(time_day, na.rm = TRUE)),
    color = "blue",
    linetype = "dashed",
    size = 1
  ) +
  labs(
    title = "Trial Duration Distribution",
    subtitle = "Red = Mean, Blue = Median",
    x = "Duration (Days)",
    y = "Count"
  ) +
  theme_minimal(base_size = 14)
```

##Distribution of Enrollment
```{r}
ggplot(train_filtered, aes(x = enrollment)) +
  geom_histogram(
    binwidth = 50,          # 每100一个区间
    fill = "#4C78A8",
    color = "white",
    alpha = 0.9
  ) +
  coord_cartesian(xlim = c(0, 2000)) +   # 👈 关键：只展示主要部分
  labs(
    title = "Enrollment Distribution",
    subtitle = "Bin width = 50",
    x = "Enrollment",
    y = "Count"
  ) +
  theme_minimal(base_size = 14)
```


##Heatamp
```{r}
importance_df <- metrics_df %>%
  group_by(model) %>%
  mutate(
    baseline_MAE = MAE[feature_setting == "all_features"]
  ) %>%
  filter(feature_setting != "all_features") %>%
  mutate(
    delta_MAE = MAE - baseline_MAE
  )

importance_wide <- importance_df %>%
  select(model, feature_setting, delta_MAE) %>%
  tidyr::pivot_wider(
    names_from = model,
    values_from = delta_MAE
  )

importance_avg <- importance_df %>%
  group_by(feature_setting) %>%
  summarise(
    mean_delta = mean(delta_MAE),
    sd_delta = sd(delta_MAE)
  ) %>%
  arrange(desc(mean_delta))
ggplot(importance_df,
       aes(x = model, y = feature_setting, fill = delta_MAE)) +
  geom_tile() +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red") +
  labs(
    title = "Feature Importance Heatmap",
    x = "Model",
    y = "Feature Removed",
    fill = "ΔMAE"
  ) +
  theme_bw()
```

##Table of outliers
```{r}
outlier<- read.csv("original_vs_no_extreme_summary.csv")
outlier_df <- outlier %>%
  filter(split == "test") %>%
  select(case, MAE, RMSE, R2, Pearson_r)

outlier_df
```




















